Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-03-02 jso.qmd
revising section 6.1

# Assessing optimisers: two simulation studies {#sec-sim-deets}

@@

@@

Performance of JSO is assessed based on the final projections across the
two optimisers: JSO and CRS. The final projections found by the two
optimisers are presented in @fig-proj, broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50% of cases. Compared to CRS, JSO achieves higher index values and
clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.

## Factors affecting JSO success rate: JSO hyper-parameters and index properties

@@

@@

```{r}
#| fig.height: 7.5
#| fig.width: 10
#| label: fig-proj
#| fig.align: "center"
#| fig.cap: "Projections found by the JSO and CRS at each 10th quantile across 50 simulations. The projection pursuit problem is to find the pipe shape using the holes index in the 6, 8, 10, and 12-dimensional spaces. The JSO uses 100 jellyfish and a maximum number of tries of 100. The CRS uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-D data space, JSO always finds a clear pipe shape while the CRS also finds the pipe shape but with a wide rim. At higher data dimensions, JSO finds a higher index value and a clearer pipe shape across all the quantiles than the CRS"

aaa <- map_dfr(
  list(pipe1000, pipe1000_8d, pipe1000_10d, pipe1000_12d), 
  ~{
    d_dim = dim(.x)[2]
    pipe_jellyfish |> 
      filter(d == d_dim) |> 
      arrange(index_val) |>
      filter(row_number() %in% c(1, seq(5, 50, 5))) |> 
      mutate(id = factor(
        paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th"),
        levels = paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th"))) |> 
      compute_projection(data = .x, col = c("d", "index_val", "id"))
  }
)

p1 <- aaa |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = aaa |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

bbb <- pipe_better |> 
  group_by(dim) |> 
  arrange(index_val) |> 
  filter(row_number() %in% c(1, seq(5, 50, 5))) |>  
  mutate(id = factor(
    paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th"),
    levels = paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th")), 
    d = 4 + 2 * as.numeric(dim)) |> 
  unnest(proj)

p2 <- bbb |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = bbb |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

>>>>

# Assessing the optimisers {#sec-sim-deets}

@@

This section summarises the findings from the simulations that compare the JSO performance with the existing CRS optimiser, and the relationship between optimisation success and hyper-parameter choices and PP index properties.

```{r}
#| fig.height: 7.5
#| fig.width: 10
#| label: fig-proj
#| fig.align: "center"
#| fig.cap: "Visual comparison of JSO and CRS results, using optimal data projections obtained over 50 simulations of the pipe data. Rows correspond to data dimension. Columns correspond to quantiles of the index values, with 0 being minimum, 100 being maximum, and 50 the median. JSO achieves the better views of the pipe generally than CRS. As dimension increases both have more difficulty finding the pipe."

aaa <- map_dfr(
  list(pipe1000, pipe1000_8d, pipe1000_10d, pipe1000_12d), 
  ~{
    d_dim = dim(.x)[2]
    pipe_jellyfish |> 
      filter(d == d_dim) |> 
      arrange(index_val) |>
      filter(row_number() %in% c(1, seq(5, 50, 5))) |> 
      mutate(id = factor(
        paste0(c(0, seq(5, 50, 5)) / 50 * 100),
        levels = c(0, seq(5, 50, 5) / 50 * 100))) |> 
      compute_projection(data = .x, col = c("d", "index_val", "id"))
  }
)

p1 <- aaa |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = aaa |> select(-V1, -V2) |> unique(), 
            aes(label = sprintf("%.3f", index_val)),
            x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

bbb <- pipe_better |> 
  group_by(dim) |> 
  arrange(index_val) |> 
  filter(row_number() %in% c(1, seq(5, 50, 5))) |>  
  mutate(id = factor(
    paste0(c(0, seq(5, 50, 5)) / 50 * 100),
        levels = c(0, seq(5, 50, 5) / 50 * 100)), 
    d = 4 + 2 * as.numeric(dim)) |> 
  unnest(proj)

p2 <- bbb |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = bbb |> select(-V1, -V2) |> unique(), 
            aes(label = sprintf("%.3f", index_val)), x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

@@

The performance of JSO is assessed relative to the existing CRS optimiser based on the optimal projection obtained. The final projections for the pipe data sets found by the two optimisers are shown in @fig-proj, with rows corresponding to data dimension. The 50 simulations in each data dimension are sorted by index value, and the projections corresponding to each 10th quantile value are shown, with minimum at left and maximum at right. Index value is printed on the plot. The purpose is to summarise the views of the data resulting from different optimisations, and hence compare results between the two optimisers. Generally, the JSO does a more consistent job of finding the pipe structure clearly. As the data dimension increases, both optimisers struggle and less clearly capture the circle shape.

<!-- broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50% of cases. Compared to CRS, JSO achieves higher index values and
clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.-->

## Effect of hyper-parameters effect on JSO success rate

@@

## Effect of index properties on JSO success rate

@@

### 2025-03-02 jso.tex
revising section 6.1

\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, colframe=quarto-callout-note-color-frame, titlerule=0mm, opacityback=0, bottomrule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, colback=white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black]

@@

\section{Assessing optimisers: two simulation
studies}\label{sec-sim-deets}

@@

Performance of JSO is assessed based on the final projections across the
two optimisers: JSO and CRS. The final projections found by the two
optimisers are presented in Figure~\ref{fig-proj}, broken down by 10th
quantile, faceted by the data dimensions. In the 6-dimensional data
scenario, JSO consistently identifies a clear pipe shape. The CRS also
finds the pipe shape but with a wide rim, suggesting a further polish
search may be required. With increasing dimensions, JSO may not always
identify the pipe shape due to random sampling, but it still finds the
pipe shape in over 50\% of cases. Compared to CRS, JSO achieves higher
index values and clearer pipe shapes across all quantiles in data of 8,
10, 12 dimensions, suggesting its advantage in exploring
high-dimensional spaces.

\subsection{Factors affecting JSO success rate: JSO hyper-parameters and
index
properties}\label{factors-affecting-jso-success-rate-jso-hyper-parameters-and-index-properties}

@@

\caption{\label{fig-proj}Projections found by the JSO and CRS at each
10th quantile across 50 simulations. The projection pursuit problem is
to find the pipe shape using the holes index in the 6, 8, 10, and
12-dimensional spaces. The JSO uses 100 jellyfish and a maximum number
of tries of 100. The CRS uses a maximum of 1000 tries in each step of
random sampling step before the algorithm terminates. In the 6-D data
space, JSO always finds a clear pipe shape while the CRS also finds the
pipe shape but with a wide rim. At higher data dimensions, JSO finds a
higher index value and a clearer pipe shape across all the quantiles
than the CRS}

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, opacityback=0, coltitle=black, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, breakable, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, rightrule=.15mm, titlerule=0mm]

@@

\section{Assessing the optimisers}\label{sec-sim-deets}

@@

This section summarises the findings from the simulations that compare
the JSO performance with the existing CRS optimiser, and the
relationship between optimisation success and hyper-parameter choices
and PP index properties.

@@

\caption{\label{fig-proj}Visual comparison of JSO and CRS results, using
optimal data projections obtained over 50 simulations of the pipe data.
Rows correspond to data dimension. Columns correspond to quantiles of
the index values, with 0 being minimum, 100 being maximum, and 50 the
median. JSO achieves the better views of the pipe generally than CRS. As
dimension increases both have more difficulty finding the pipe.}

@@

The performance of JSO is assessed relative to the existing CRS
optimiser based on the optimal projection obtained. The final
projections for the pipe data sets found by the two optimisers are shown
in Figure~\ref{fig-proj}, with rows corresponding to data dimension. The
50 simulations in each data dimension are sorted by index value, and the
projections corresponding to each 10th quantile value are shown, with
minimum at left and maximum at right. Index value is printed on the
plot. The purpose is to summarise the views of the data resulting from
different optimisations, and hence compare results between the two
optimisers. Generally, the JSO does a more consistent job of finding the
pipe structure clearly. As the data dimension increases, both optimisers
struggle and less clearly capture the circle shape.

\subsection{Effect of hyper-parameters effect on JSO success
rate}\label{effect-of-hyper-parameters-effect-on-jso-success-rate}

\subsection{Effect of index properties on JSO success
rate}\label{effect-of-index-properties-on-jso-success-rate}
